Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-09-01 jso.qmd
revised the introduction

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

```{=html}
<!--  
\begin{assumption}\label{assum:my_assumption}
This is an assumption.
\end{assumption}
Assumption \ref{assum:my_assumption} is convenient.

@@

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
-->
```
Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising an objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Let $X \in \mathbb{R}^{n\times p}$ be the data matrix, $A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix, where $A$ belongs to the Stiefel manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. The projection $Y = XA$ is a linear transformation that maps data from a $p$-dimensional space into a $d$-dimensional space. The index function $f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ is a scalar function that measures an interesting aspect of the projected data, such as deviation from normality, presence of clusters, non-linear structure, or other features of interest. For a fixed data, PP finds the orthonormal basis $A$ that maximises the index value of the projection, $Y = XA$:

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieve a globally optimal viewpoint of the data.

Optimisation of PP is typically investigated in the literature when new
indexes are proposed [@posse1995; @marie-sainte2010; @grochowski2011] or
when visualisation method are used to track the optimisation process.
@cook1995grand introduced the PP guided tour, which monitors the
optimisation visually to see the data projections leading in and out of
the optima. An implementation is available in the `tourr` package
[@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose
optimisation processes, particularly focusing on the guided tour, and
revealed a need for improved optimisation. While improving the quality
of the optimisation solutions in the tour is essential, it is also
important to be able to view the data projections as the optimisation
progresses. Integrating the guided tour with a global optimisation
algorithm that is efficient in finding the global optimal and enables
viewing of the projected data during the exploration process is a goal.

>>>>

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(tourr)
library(PPtreeViz)
library(ferrn)
library(colorspace)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

<!--  
```{=html}
\begin{assumption}\label{assum:my_assumption}
This is an assumption.
\end{assumption}
Assumption \ref{assum:my_assumption} is convenient.

@@

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
```
-->

Projection Pursuit (PP) (@kr69, @FT74, @huber85) is a dimension reduction technique aimed at identifying informative linear projections of data. This is useful for exploring high-dimensional data, and creating plots of the data for publication that reveal the main features.  The method involves optimising an objective function known as the PP index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the criteria for what constitutes interesting or informative projections. Let $X \in \mathbb{R}^{n\times p}$ be the data matrix, $A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix, where $A$ belongs to the Stiefel manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. (It is interesting to note that when using PP visually, one cares less about $A$ than the plane described by $A$, because the orientation in the plane is irrelevant. The the space of planes belongs to a Grassmann manifold.) The projection $Y = XA$ is a linear transformation that maps data from a $p$-dimensional space into a $d$-dimensional space. The index function $f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ is a scalar function that measures an interesting aspect of the projected data, such as deviation from normality, presence of clusters, non-linear structure, or other features of interest. For a fixed data, PP finds the orthonormal basis $A$ that maximises the index value of the projection, $Y = XA$:

@@

The index functions vary in structure, partly depending on the data that is being projected. @fig-example-functions shows two examples. Huber plots [@huberplot] of 2D data sets are in (a) and (c). The PP index values for all 1D projections of the 2D data are shown in polar coordinates, revealing the nature of the function. The dashed circle is a baseline set at the average value, and the straight line marks the optimal projection. Plots (b) and (d) show the respective best projections of the data as histograms. Indexes like the holes, central mass and skewness [@cook1993projection] are generally smooth for most data sets. Many indexes noisy and non-convex, requiring an effective and efficient optimisation procedure to explore the data landscape and achieve a globally optimal viewpoint of the data. The skewness index computed for trimodal data, in (a), is smooth with a large squint angle but has three modes, and thus is not convex. The binned normality index computed on the famous RANDU data, in (c), is noisier and has a very small squint angle. The discreteness cannot be seen unless the optimiser is very close to the optimal projection.

```{r}
#| label: huber-plot-function
#| echo: false
huber_plot <- function (origdata2D, origclass=NULL, 
  PPmethod = "LDA", weight = TRUE, r = 1, lambda = 0.5, 
  opt.proj = TRUE, UserDefFtn = NULL, ...) 
{
    index <- NULL
    best.proj <- NULL
    best.index <- 0
    origdata2D <- as.matrix(origdata2D)
    for (i in 0:360) {
        theta <- pi/180 * i
        proj.data <- matrix(cos(theta) * origdata2D[, 1] + sin(theta) * 
            origdata2D[, 2])
        proj <- matrix(c(cos(theta), sin(theta)), ncol = 1)
        if (PPmethod == "LDA") {
            newindex <- PPtreeViz::LDAindex(origclass, origdata2D, proj = proj, 
                weight = weight)
        }
        else if (PPmethod == "PDA") {
            newindex <- PPtreeViz::PDAindex(origclass, origdata2D, proj, 
                weight = weight, lambda = lambda)
        }
        else if (PPmethod == "Lr") {
            newindex <- PPtreeViz::Lrindex(origclass, origdata2D, proj, 
                weight = weight, r = r)
        }
        else if (PPmethod == "GINI") {
            newindex <- PPtreeViz::GINIindex1D(origclass, origdata2D, proj)
        }
        else if (PPmethod == "ENTROPY") {
            newindex <- PPtreeViz::ENTROPYindex1D(origclass, origdata2D, 
                proj)
        }
        else if (PPmethod == "holes") {
          newindex <- tourr::holes()(proj.data)
        }
        else if (PPmethod == "skewness") {
          newindex <- tourr::skewness()(proj.data)
        }
        else if (PPmethod == "norm_bin") {
          newindex <- tourr::norm_bin(nrow(proj.data))(proj.data)
        }
        else if (PPmethod == "UserDef") {
            newindex <- UserDefFtn(proj.data, ...)
        }
        index <- c(index, newindex)
    }
    sel.index <- which(index[1:360] > signif(max(index), 6) - 
        1e-06)
    theta.best.all <- pi/180 * (sel.index - 1)
    theta.best <- theta.best.all[1]
    proj.data.best <- matrix(cos(theta.best) * origdata2D[, 1] + 
        sin(theta.best) * origdata2D[, 2])
    index.best <- max(index)
    range <- round(max(index) - min(index), 5)
    if (range == 0) {
      PPindex <- rep(4, length(index))
    }
    else {
      PPindex <- (index - min(index))/range * 2 + 3
    }
    data.circle <- NULL
    data.index <- NULL
    for (i in 1:361) {
        theta <- pi/180 * (i - 1)
        data.index <- rbind(data.index, c(PPindex[i] * cos(theta), 
            PPindex[i] * sin(theta)))
        data.circle <- rbind(data.circle, c(4 * cos(theta), 4 * 
            sin(theta)))
    }
    maxdiff <- max(c(diff(range(origdata2D[, 1])), diff(range(origdata2D[, 
        2]))))
    orig.scaled <- apply(origdata2D, 2, function(x) (x - mean(x))/maxdiff * 
        3.5)
    data.cX <- data.circle[, 1]
    data.cY <- data.circle[, 2]
    data.X <- data.index[, 1]
    data.Y <- data.index[, 2]
    plot.data <- data.frame(data.cX, data.cY, data.X, data.Y)
    x <- orig.scaled[, 1]
    y <- orig.scaled[, 2]
    if (is.null(origclass)) 
      group <- rep(1, nrow(origdata2D))
    else
      group <- origclass
    point.data <- data.frame(x, y, group)
    min.X <- min(unlist(plot.data))
    max.X <- max(unlist(plot.data))
    if (is.null(origclass)) {
      P1 <- ggplot(data = plot.data, aes(x = data.X, y = data.Y)) + 
        geom_path() + geom_path(aes(x = data.cX, y = data.cY), 
        linetype = "dashed") + geom_point(data = point.data, 
        aes(x = x, y = y)) + scale_x_continuous(breaks = NULL) + 
        scale_y_continuous(breaks = NULL) + xlab("") + ylab("") + 
        coord_fixed() + theme_bw() + theme(panel.border = element_blank())
    } else {
      P1 <- ggplot(data = plot.data, aes(x = data.X, y = data.Y)) + 
        geom_path() + geom_path(aes(x = data.cX, y = data.cY), 
        linetype = "dashed") + geom_point(data = point.data, 
        aes(x = x, y = y, color = group, shape = group)) +
        scale_x_continuous(breaks = NULL) + 
        scale_y_continuous(breaks = NULL) + xlab("") + ylab("") + 
        coord_fixed() + theme_bw() + theme(panel.border = element_blank())
    }
    if (opt.proj) {
        P1 <- P1 + geom_abline(intercept = 0, slope = sin(theta.best)/cos(theta.best))
        if (length(theta.best.all) > 1) 
            for (i in 2:length(theta.best.all)) P1 <- P1 + geom_abline(intercept = 0, 
                slope = sin(theta.best.all[i])/cos(theta.best.all[i]), 
                linetype = "dashed")
    }
    best.proj.data <- proj.data.best
    #group <- origclass
    if (is.null(origclass)) {
      hist.data <- data.frame(best.proj.data)
      P2 <- ggplot(data = hist.data, aes(x = best.proj.data)) + 
        geom_histogram()
    } else {
      hist.data <- data.frame(best.proj.data, group)
      P2 <- ggplot(data = hist.data, aes(x = best.proj.data, 
                                         group = group)) + 
        geom_histogram(aes(fill = group), position = "stack")
    }
    #gridExtra::grid.arrange(P1, P2, nrow = 1)
    return(list(plot=P1, proj_data=best.proj.data))
}

```{r}
#| label: fig-example-functions
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 80%
#| fig-cap: "Examples of PP indexes with large (top row) and small (bottom row) squint angles, shown with a Huber plot, and histogram of the projected data corresponding to the optimal projection. Huber plot shows PP index values for all 1D projections in polar coordinates."
data(randu)
randu_std <- as.data.frame(apply(randu, 2, function(x) (x-mean(x))/sd(x)))
randu_std$yz <- sqrt(35)/6*randu_std$y-randu_std$z/6

#proj <- animate_xy(flea[,1:6], guided_tour(lda_pp(flea$species)))
best_proj <- matrix(c(-0.44157005,  0.80959831,
                      0.14622089, -0.08228193,
                      -0.08454314, -0.23234287,
                      0.70367852,  0.28185021,
                      0.27876825,  0.44819691,
                      0.45123453,  0.05896646), ncol=2, byrow=T)
flea2D <- as.matrix(flea[,1:6]) %*% best_proj
colnames(flea2D) <- c("X1", "X2")
flea2D <- as.data.frame(flea2D)

# Make both plots
large_sqa <- huber_plot(flea2D, PPmethod="skewness")
small_sqa <- huber_plot(randu_std[c(1,4)], PPmethod="norm_bin")

sqa1 <- large_sqa$plot + ggtitle("(a) Skewness index")
lsqa_d <- tibble(d = large_sqa$proj_data[,1])
sqa2 <- ggplot(lsqa_d, aes(x=d)) + 
  geom_histogram(binwidth=0.25, colour="white") +
  xlab("") + ylab("") +
  ggtitle("(b) Optimally projected data") +
  theme_bw() +
  theme(axis.text.y = element_blank())
sqa3 <- small_sqa$plot + ggtitle("(c) Binned normality index")
ssqa_d <- tibble(d = small_sqa$proj_data[,1])
sqa4 = ggplot(ssqa_d, aes(x=d)) + 
  geom_histogram(breaks = seq(-2.2, 2.4, 0.12)) +
  xlab("") + ylab("") +
  ggtitle("(d) Optimally projected data") +
  theme_bw() +
  theme(axis.text.y = element_blank())

sqa1 + sqa2 + sqa3 + sqa4 + plot_layout(ncol=2, widths=c(2,2))
```

Optimisation of PP is often discussed when new indexes are proposed [@posse1995; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the optimisation more closely to the index, when they introduced the PP guided tour, which monitors the optimisation visually so that the user can see the projected data leading in and out of the optima. An implementation is available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose optimisation processes, particularly focusing on the guided tour, and revealed a need for improved optimisation. While improving the quality of the optimisation solutions in the tour is essential, it is also important to be able to view the data projections as the optimisation progresses. Integrating the guided tour with a global optimisation algorithm that is efficient in finding the global optimal and enables viewing of the projected data during the exploration process is a goal.

### 2024-09-01 jso.tex
revised the introduction

Projection Pursuit (PP) (Kruskal (1969), Friedman and Tukey (1974)) is a
dimension reduction technique aimed at identifying informative linear
projections of data. The method involves optimising an objective
function known as the PP index (e.g. Hall (1989), D. Cook, Buja, and
Cabrera (1993), Lee and Cook (2010), Loperfido (2018), Loperfido
(2020)), which defines the criteria for what constitutes interesting or
informative projections. Let \(X \in \mathbb{R}^{n\times p}\) be the
data matrix, \(A \in\mathbb{R}^{p \times d}\) be an orthonormal matrix,
where \(A\) belongs to the Stiefel manifold
\(\mathcal{A} = V_d(\mathbb{R}^p)\). The projection \(Y = XA\) is a
linear transformation that maps data from a \(p\)-dimensional space into
a \(d\)-dimensional space. The index function
\(f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}\) is a scalar function
that measures an interesting aspect of the projected data, such as
deviation from normality, presence of clusters, non-linear structure, or
other features of interest. For a fixed data, PP finds the orthonormal
basis \(A\) that maximises the index value of the projection,
\(Y = XA\):

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieve a globally optimal viewpoint of the data.

Optimisation of PP is typically investigated in the literature when new
indexes are proposed (Posse 1995a; Marie-Sainte, Berro, and Ruiz-Gazen
2010; Grochowski and Duch 2011) or when visualisation method are used to
track the optimisation process. Dianne Cook et al. (1995) introduced the
PP guided tour, which monitors the optimisation visually to see the data
projections leading in and out of the optima. An implementation is
available in the \texttt{tourr} package (Wickham et al. 2011) in R (R
Core Team 2023). Zhang et al. (2021) illustrated how to diagnose
optimisation processes, particularly focusing on the guided tour, and
revealed a need for improved optimisation. While improving the quality
of the optimisation solutions in the tour is essential, it is also
important to be able to view the data projections as the optimisation
progresses. Integrating the guided tour with a global optimisation
algorithm that is efficient in finding the global optimal and enables
viewing of the projected data during the exploration process is a goal.

@@

There are several PP index functions available: \texttt{holes()} and
\texttt{cmass()} (D. Cook, Buja, and Cabrera 1993); \texttt{lda\_pp()}
(Lee et al. 2005); \texttt{pda\_pp()} (Lee and Cook 2010);
\texttt{dcor2d()} and \texttt{splines2d()} (Grimm 2016);
\texttt{norm\_bin()} and \texttt{norm\_kol()} (Huber 1985);
\texttt{slice\_index()} (Ursula Laa and Valencia 2022). Most are
relatively simply defined, for any projection dimension, and implemented
because they are relatively easy to optimise. A goal is to be able to
incorporate more complex PP indexes, for example, based on scagnostics
(L. Wilkinson, Anand, and Grossman (2005), Leland Wilkinson and Wills
(2008)).

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, breakable, arc=.35mm, opacityback=0, toprule=.15mm, opacitybacktitle=0.6, left=2mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, bottomrule=.15mm]

@@

\bibitem[\citeproctext]{ref-cook1995grand}
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
{``Grand Tour and Projection Pursuit.''} \emph{Journal of Computational
and Graphical Statistics} 4 (3): 155--72.
\url{https://doi.org/10.1080/10618600.1995.10474674}.

@@

>>>>

Projection Pursuit (PP) (Kruskal (1969), Friedman and Tukey (1974),
Huber (1985)) is a dimension reduction technique aimed at identifying
informative linear projections of data. This is useful for exploring
high-dimensional data, and creating plots of the data for publication
that reveal the main features. The method involves optimising an
objective function known as the PP index (e.g. Hall (1989), Cook, Buja,
and Cabrera (1993), Lee and Cook (2010), Loperfido (2018), Loperfido
(2020)), which defines the criteria for what constitutes interesting or
informative projections. Let \(X \in \mathbb{R}^{n\times p}\) be the
data matrix, \(A \in\mathbb{R}^{p \times d}\) be an orthonormal matrix,
where \(A\) belongs to the Stiefel manifold
\(\mathcal{A} = V_d(\mathbb{R}^p)\). (It is interesting to note that
when using PP visually, one cares less about \(A\) than the plane
described by \(A\), because the orientation in the plane is irrelevant.
The the space of planes belongs to a Grassmann manifold.) The projection
\(Y = XA\) is a linear transformation that maps data from a
\(p\)-dimensional space into a \(d\)-dimensional space. The index
function \(f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}\) is a scalar
function that measures an interesting aspect of the projected data, such
as deviation from normality, presence of clusters, non-linear structure,
or other features of interest. For a fixed data, PP finds the
orthonormal basis \(A\) that maximises the index value of the
projection, \(Y = XA\):

@@

The index functions vary in structure, partly depending on the data that
is being projected. Figure~\ref{fig-example-functions} shows two
examples. Huber plots (Huber 1990) of 2D data sets are in (a) and (c).
The PP index values for all 1D projections of the 2D data are shown in
polar coordinates, revealing the nature of the function. The dashed
circle is a baseline set at the average value, and the straight line
marks the optimal projection. Plots (b) and (d) show the respective best
projections of the data as histograms. Indexes like the holes, central
mass and skewness (Cook, Buja, and Cabrera 1993) are generally smooth
for most data sets. Many indexes noisy and non-convex, requiring an
effective and efficient optimisation procedure to explore the data
landscape and achieve a globally optimal viewpoint of the data. The
skewness index computed for trimodal data, in (a), is smooth with a
large squint angle but has three modes, and thus is not convex. The
binned normality index computed on the famous RANDU data, in (c), is
noisier and has a very small squint angle. The discreteness cannot be
seen unless the optimiser is very close to the optimal projection.

\includegraphics[width=0.8\textwidth,height=\textheight]{jso_files/figure-pdf/fig-example-functions-1.pdf}

\caption{\label{fig-example-functions}Examples of PP indexes with large
(top row) and small (bottom row) squint angles, shown with a Huber plot,
and histogram of the projected data corresponding to the optimal
projection. Huber plot shows PP index values for all 1D projections in
polar coordinates.}

Optimisation of PP is often discussed when new indexes are proposed
(Posse 1995a; Marie-Sainte, Berro, and Ruiz-Gazen 2010; Grochowski and
Duch 2011). Cook et al. (1995) tied the optimisation more closely to the
index, when they introduced the PP guided tour, which monitors the
optimisation visually so that the user can see the projected data
leading in and out of the optima. An implementation is available in the
\texttt{tourr} package (Wickham et al. 2011) in R (R Core Team 2023).
Zhang et al. (2021) illustrated how to diagnose optimisation processes,
particularly focusing on the guided tour, and revealed a need for
improved optimisation. While improving the quality of the optimisation
solutions in the tour is essential, it is also important to be able to
view the data projections as the optimisation progresses. Integrating
the guided tour with a global optimisation algorithm that is efficient
in finding the global optimal and enables viewing of the projected data
during the exploration process is a goal.

@@

There are several PP index functions available: \texttt{holes()} and
\texttt{cmass()} (Cook, Buja, and Cabrera 1993); \texttt{lda\_pp()} (Lee
et al. 2005); \texttt{pda\_pp()} (Lee and Cook 2010); \texttt{dcor2d()}
and \texttt{splines2d()} (Grimm 2016); \texttt{norm\_bin()} and
\texttt{norm\_kol()} (Huber 1985); \texttt{slice\_index()} (Ursula Laa
and Valencia 2022). Most are relatively simply defined, for any
projection dimension, and implemented because they are relatively easy
to optimise. A goal is to be able to incorporate more complex PP
indexes, for example, based on scagnostics (L. Wilkinson, Anand, and
Grossman (2005), Leland Wilkinson and Wills (2008)).

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, toptitle=1mm, rightrule=.15mm, colback=white, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, arc=.35mm, bottomrule=.15mm, toprule=.15mm, titlerule=0mm, bottomtitle=1mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, breakable, opacitybacktitle=0.6]

@@

\bibitem[\citeproctext]{ref-cook1995grand}
Cook, D., A. Buja, J. Cabrera, and C. Hurley. 1995. {``Grand Tour and
Projection Pursuit.''} \emph{Journal of Computational and Graphical
Statistics} 4 (3): 155--72.
\url{https://doi.org/10.1080/10618600.1995.10474674}.

@@

\bibitem[\citeproctext]{ref-huberplot}
---------. 1990. {``Data Analysis and Projection Pursuit.''} Technical
Report PJH-90-1. Dept. of Mathematics, Massachusetts Institute of
Technology.
